Codebook
========
Codebook was generated on `r as.character(Sys.time())` during the same process that generated the dataset. See `run_analysis.md` or `run_analysis.html` for details on dataset creation.

Variable list and descriptions
------------------------------

Variable name | Description
-----------------|------------
subject | ID the subject who performed the activity for each window sample. Its range is from 1 to 30.
activity | Activity name
featDomain | Feature: Time domain signal or frequency domain signal (Time or Freq)
featInstrument | Feature: Measuring instrument (Accelerometer or Gyroscope)
featAcceleration | Feature: Acceleration signal (Body or Gravity)
featVariable | Feature: Variable (Mean or SD)
featJerk | Feature: Jerk signal
featMagnitude | Feature: Magnitude of the signals calculated using the Euclidean norm
featAxis | Feature: 3-axial signals in the X, Y and Z directions (X, Y, or Z)
featCount | Feature: Count of data points used to compute `average`
featAverage | Feature: Average of each variable for each activity and each subject

DataSet Structure
-----------------
```{r}
str(dataDataTableTidy)
```

Data Table Key Variable Listing
-------------------------------
```{r}
key(dataDataTableTidy)
```

DataSet Preview
---------------
```{r}
dataDataTableTidy
```

Variable Summary
----------------
```{r}
summary(dataDataTableTidy)
```

Feature Combination Listing
---------------------------
```{r}
dataDataTableTidy[, 
                  .N, 
                  by=c(names(dataDataTableTidy)[grep("^feat", names(dataDataTableTidy))])]
```

Tidy Data Table File Persistence
--------------------------------
**Save tidy data table to tab-delimited `HumanActivityRecognitionDataSet.txt` text file.**
```{r SaveTidyDataTable}
outputFile <- file.path(path, "HumanActivityRecognitionDataSet.txt")

write.table(dataDataTableTidy, 
            outputFile, 
            quote=FALSE, 
            sep="\t", 
            row.names=FALSE)
```




